---
title: "Whole Genome Coverage"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  prettydoc::html_pretty:
    theme: cayman
    highlight: vignette
    toc: true
    fig_caption: yes
editor_options:
  chunk_output_type: console
---

```{r filepaths, echo=F, warning=F, message=F, results='hide'}
alignsumfilepath = "/pine/o/n/onyen/SomFoler/SumSTATsandQC/AlignSummary/"
flagsumfilepath = "/pine/o/n/onyen/SomFoler/SumSTATsandQC/FlagStats/"
calllocisumfilepath = "/pine/o/n/onyen/SomFoler/SumSTATsandQC/CallableLoci/"
gencovdirdatapath = "/pine/o/n/onyen/SomFoler/SumSTATsandQC/coverage/data/"

```

```{r imports, echo=F, warning=F, message=F, results='hide'}
library(tidyverse)
library(RColorBrewer)
library(devtools)
devtools::install_github("nickbrazeau/NFBtools")
library(NFBtools)
library(DT)
options(bitmapType='cairo')
```

# Background
words

## Goal
**Overall**: words
**Immediate**: words


## Alignmment Summary
```{r alignsummary, echo=F, warning=F, message=F, results='asis'}
alignsumfiles <- dir(path=alignsumfilepath, full.names = TRUE)

readalign <- function(path){
  out <- read_tsv(path,  skip=(grep("CATEGORY", readLines(alignsumfiles[1]))-1))
  sample <- stringr::str_split_fixed(string=basename(path), pattern="_", n=2)[1]
  out <- as.data.frame(cbind(sample, out))
  return(out)
}

alignsumfiles <- lapply(alignsumfiles, readalign)
alignsumfiles <- do.call("rbind", alignsumfiles)

DT::datatable(alignsumfiles, extensions = 'Buttons', options = list(dom = 'Bfrtip',buttons = c('csv')))


```


## FlagStats
```{r flagstat, echo=F, warning=F, message=F, results='asis'}
flagsumfiles <- dir(path=flagsumfilepath, full.names = TRUE)

readflag <- function(path){
  out <- read_tsv(path, col_names = F)
  sample <- stringr::str_split_fixed(string=basename(path), pattern="_", n=2)[1]
  out <- as.data.frame(cbind(sample, out))
  colnames(out) <- c("sample", "summary_stat")
  return(out)
}

flagsumfiles <- lapply(flagsumfiles, readflag)
flagsumfiles <- do.call("rbind", flagsumfiles)

DT::datatable(flagsumfiles, extensions = 'Buttons', options = list(dom = 'Bfrtip',buttons = c('csv')))


```


## Callable Loci Summary
Output is based on a mimimum base quality of 20, minimum mapping quality of 10, and a minimum depht of 4 (these are all > criterion).  

```{r calloci, echo=F, warning=F, message=F, results='asis'}
calllocisumfiles <- dir(path=calllocisumfilepath, full.names = TRUE)
calllocisumfiles <- calllocisumfiles[grepl(pattern = "_summarytable.txt", x = basename(calllocisumfiles))]
readcallloci <- function(path){
  out <- read_tsv(path, col_names = T)
  sample <- stringr::str_split_fixed(string=basename(path), pattern="_", n=2)[1]
  out <- as.data.frame(cbind(sample, out))
  return(out)
}

calllocisumfiles <- lapply(calllocisumfiles, readcallloci)
calllocisumfiles <- do.call("rbind", calllocisumfiles)

DT::datatable(calllocisumfiles, extensions = 'Buttons', options = list(dom = 'Bfrtip',buttons = c('csv')))


```

## Genomic Coverage by Sample
Presented is the overall Genomic Coverage and a Summary Table by Sample.
```{r genomcov, echo=F, warning=F, message=F, fig.width=12, fig.height=12, dpi=500}

#------------------------
# read in
#------------------------
bamcovs <- NFBtools::bedtoolsgenomecov2bamCov(gencovdir = gencovdirdatapath)

#------------------------
# genomic summary plot
#------------------------
genomcovplot <- NFBtools::bamCov2OverallPercCov(bamcovs)


#------------------------
# genomic summary table
#------------------------
genomsummarydepth <- do.call("rbind", lapply(bamcovs, function(x){x[["genomsummarydepth"]]}))
#https://stackoverflow.com/questions/11774703/adding-text-to-a-grid-table-plot/11775211#11775211
genomtable <- gridExtra::tableGrob(genomsummarydepth, rows = NULL)
title <- grid::textGrob("Summary of Coverage Depth by Sample", gp=grid::gpar(fontfamily="Arial", fontsize=14, fontface="bold"))
padding <- unit(5,"mm")
genomtable <- gtable::gtable_add_rows(
        genomtable,
        heights = grid::grobHeight(title) + padding,
        pos = 0)
genomtable <- gtable::gtable_add_grob(
        genomtable,
        list(title),
        1, 1, 1, ncol(genomtable))

grobs = list(genomcovplot, genomtable)
gridExtra::grid.arrange(grobs = grobs, layout_matrix = rbind(c(1),
                                                             c(1),
                                                             c(2)
                                                             )
                        )



```

## Sample-Level Genomic Coverage
```{r smplcov, echo=F, warning=F, message=F, fig.width=18, fig.height=12, results='hide', dpi=500}

NFBtools::bamCov2SmplRaster(input = bamcovs)

```

**This report was generated using NFBtools by Nick Brazeau (please see github and NFBtools to customize further)**.
